---
title: "732A90 Lab 3"
output:
  pdf_document: default
  html_document: default
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Computational Statistics - Fall 23

# Computer Lab 3

### Hugo Morvan, Daniele Bozzoli

## Question 1: Sampling algorithms for a triangle distribution

f(x) = 
0         if x < −1 or x > 1,
x + 1     if − 1 ≤ x ≤ 0,
1 − x     if 0 < x ≤ 1.

### a.
Choose an appropriate and simple envelope e(x) for the density and program a random generator for X using rejection sampling.

```{r a}

```

### b.
In Lecture 3, another triangle distribution was generated using the inverse cumulative distribution function method, see page 9-10 of the lecture notes. Let Y be a random variable following this distribution. A random variable −Y has a triangle distribution on the interval [−1, 0]. Program a random generator for X using composition sampling based on Y and −Y . You can use the code from the lecture to generate Y .

```{r b}

```

### c.
Sums or differences of two independent uniformly distributed variables can also have some triangle distribution. When U1 , U2 are two independent U nif [0, 1] random variables, U1 − U2 has the same distribution as X. Use this result to program a generator for X.

```{r}

```

### d.
Check your random generators in each of a. to c. by generating 10000 random variables and plotting
a histogram. Which of the three methods do you prefer if you had to generate samples of X? Use
the data from one method to determine the variance of X.

```{r}

```

