---
title: "lab5"
author: "Hugo Morvan, Daniele Bozzoli"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Question 1: Hypothesis testing

In 1970, the US Congress instituted a random selection process for the military draft. All 366 possible birth dates were placed in plastic capsules in a rotating drum and were selected one by one. The first date drawn from the drum received draft number one, the second date drawn received draft number two, etc. Then, eligible men were drafted in the order given by the draft number of their birth date. In a truly random lottery there should be no relationship between the date and the draft number. Your task is to investigate whether there can be doubts concerning the randomness of the selection of the draft numbers. The draft numbers (Y=Draft_No) sorted by day of year (X=Day_of_year) are given in the file lottery.csv. The data was originally published by the U.S. Government, and most conveniently made available online at <http://jse.amstat.org/jse_data_archive.htm> (see also Starr Norton (1997) Nonrandom Risk: The 1970 Draft Lottery, Journal of Statistics Education, 5:2, DOI: 10.1080/10691898.1997.11910534)

## 1.

Create a scatterplot of Y versus X, are any patterns visible?

```{r 1.1, echo=FALSE}
#1.1
data=read.csv("lottery.csv", header = TRUE, sep = ";")
summary(data)
plot(data$Day_of_year, data$Draft_No, xlim = c(0, 366), ylim = c(0, 366), xlab = "Day of year", ylab = "Draft No")
```

There is no clear pattern visible in this data. \## 2. Fit a curve to the data. First fit an ordinary linear model and then fit and then one using loess(). Do these curves suggest that the lottery is random? Explore how the resulting estimated curves are encoded and whether it is possible to identify which parameters are responsible for non--randomness.

```{r 1.2, echo=FALSE}
#1.2
model=lm(Draft_No ~ Day_of_year , data=data)
summary(model)
plot(data$Day_of_year, data$Draft_No, xlim = c(0, 366), ylim = c(0, 366), xlab = "Day of year", ylab = "Draft No")
abline(model, col="red")

model2=loess(Draft_No ~ Day_of_year , data=data)
summary(model2)
plot(data$Day_of_year, data$Draft_No, xlim = c(0, 366), ylim = c(0, 366), xlab = "Day of year", ylab = "Draft No")
lines(data$Day_of_year, model2$fitted, col="red")

```

## 3.

In order to check if the lottery is random, one can use various statistics. One such possibility is based on the expected responses. The fitted loess smoother provides an estimate $\hat{Y}$ as a function of X. It the lottery was random, we would expect $\hat{Y}$ to be a flat line, equaling the empirical mean of the observed responses, Y . The statistic we will consider will be $$ S = \sum_{i=1}^{n}|\hat{Y}_i-\bar{Y}| $$

If S is not close to zero, then this indicates some trend in the data, and throws suspicion on the randomness of the lottery. Estimate S's distribution through a non--parametric bootstrap, taking B = 2000 bootstrap samples. Decide if the lottery looks random, what is the p--value of the observed value of S.

```{r 1.3, echo=FALSE}
#1.3
S = sum(abs(model2$fitted-mean(data$Draft_No)))
B=2000
S_boot=rep(NA, B)
for (b in 1:B){
  data_boot=data[sample(1:nrow(data), replace = TRUE),]
  model_boot=loess(Draft_No ~ Day_of_year , data=data_boot)
  S_boot[b]=sum(abs(model_boot$fitted-mean(data_boot$Draft_No)))
}
p_value=mean(S_boot>S)
S
p_value
```

the p_value is 0.556 on 364 degrees of freedom. We can't reject the null hypothesis that the lottery is random (?)

## 4.

We will now want to investigate the power of our considered test. First based on the test statistic S, implement a function that tests the hypothesis H0 : Lottery is random versus H1 : Lottery is non--random. The function should return the value of S and its p--value, based on 2000 bootstrap samples.

```{r 1.4, echo=FALSE}
#1.4
#Permutation test:

test_hyp=function(data){
  model=loess(Draft_No ~ Day_of_year , data=data)
  S=sum(abs(model$fitted-mean(data$Draft_No))) #1: T(X) value of statistic from observed data 
  B=2000
  S_boot=rep(NA, B)
  #generate B bootstrap samples without replacement
  for (b in 1:B){
    #2: Create permutations g_1, g_B of group variable {If the number of permutations is too large, sample B randomly without replacement. E.g. generate random permutations and keep only unique ones.}
    data_boot = data
    data_boot$Day_of_year = sample(data$Day_of_year, replace = FALSE) #without replacement
    model_boot=loess(Draft_No ~ Day_of_year , data=data_boot) #fitting the model
    #3: Evaluate test statistic on each permutation
    S_boot[b] = sum(abs(model_boot$fitted-mean(data_boot$Draft_No))) 
    
  }
  p_value=mean(S_boot>=S) #4: Estimate p–value: p^ = #{T(X_gb) >= T(X)}/B
  return(list(S=S, p_value=p_value))
}

test_hyp(data)
```

## 5.

Now we will try to make a rough estimate of the power of the test constructed in Step 4 by generating more and more biased samples:

(a) Create a dataset of the same dimensions as the original data. Choose k, out of the 366, dates and assign them the end numbers of the lottery (i.e., they are not legible for the draw). The remaining 366 - k dates should have random numbers assigned (from the set {1,..., 366 - k}). The k dates should be chosen in two ways:

<!-- -->

i.  k consecutive dates,

ii. as blocks (randomly scattered) of $\lfloor k/3 \rfloor$ consecutive dates (this is of course for $k >= 3$, and if k is not divisible by 3, then some blocks can be of length $\lfloor k/3 \rfloor +1$).

\*\*\* We got stuck on this question. We are hopping to get some feedback and/or cues during the seminar. \*\*\*

```{r 1.5a, echo=FALSE}
#1.5
#(a)
#i k consecutive dates,

#Create a dataset of the same dimensions as the original data. 
data2=data

#Choose k, out of the 366, dates and assign them the end numbers of the lottery (i.e., they are not legible for the draw).
#===================
k=300
#===================
#Randomize the first 366-k dates
# data2$Draft_No[366-k:366]=sample(1:(366-k), k, replace = FALSE)
model2=loess(Draft_No ~ Day_of_year , data=data2)
plot(data2$Day_of_year, data2$Draft_No, xlim = c(0, 366), ylim = c(0, 366), xlab = "Day of year", ylab = "Draft No")

#ii. as blocks (randomly scattered) of $\lfloor k/3 \rfloor$  consecutive dates (this is of course for $k >= 3$, and if k is not divisible by 3, then some blocks can be of length $ \lfloor k/3 \rfloor +1$).

data3=data

get_lotery_numbers_with_chunks <- function(k){
  #k is the number of dates that are not legible for the draw
  # 3 cases : 
  # 1. n_chunks is an integer -> 3 normal chunks
  # 2. n_chunks is not an integer and has a remainder of 1 -> 2 normal chunks and 1 long chunk
  # 3. n_chunks is not an integer and has a remainder of 2 -> 1 normal chunk and 2 long
  
  #Case 1
  if(k%%3 == 0){
    chunk_size = floor(k/3)
    n_chunks = 3
    for (i in 1:n_chunks){
      #Sample n_chunks_normal chunks of length chunk_size
      #random starting point
      start = sample(1:366, 1, replace = TRUE)
      #sample chunk_size consecutive days
      data3$Draft_No[start:start+chunk_size-1] = sample(1:(366-k), chunk_size, replace = TRUE)
    }
    
  }
  
  # chunk_size = floor(k/3)
  # remainder = k%%chunk_size
  # n_chunks_normal = floor(k/chunk_size) - remainder
  # n_chunks_long = remainder
  # for (i in 1:n_chunks_normal){
  # #Sample n_chunks_normal chunks of lenght chunk_size
  # #random starting point
  # start = sample(1:366, 1, replace = TRUE)
  # #sample chunk_size consecutive days
  # data3$Draft_No[start:start+chunk_size-1] = sample(1:(366-k), chunk_size, replace = TRUE)
  # }
  # for (i in 1:n_chunks_long){
  #    #Sample n_chunks_long chunks of lenght chunk_size+1
  #   start = sample(1:366, 1, replace = TRUE)
  #   #sample chunk_size consecutive days
  #   data3$Draft_No[start:start+chunk_size] = sample(1:(366-k), chunk_size+1, replace = TRUE)
  # }
  # data3$Draft_No[n_chunks_normal*chunk_size+n_chunks_long*(chunk_size+1):366] = sample(1:(366-k), 366-k, replace = TRUE)

}



model3=loess(Draft_No ~ Day_of_year , data=data3)
plot(data3$Day_of_year, data3$Draft_No, xlim = c(0, 366), ylim = c(0, 366), xlab = "Day of year", ylab = "Draft No")
```

(b) For each of the Plug the two new not--completely--random datasets from item 5a into the bootstrap test with B = 2000 and note whether it was rejected.

```{r 1.5b}
#(b)

S=sum(abs(model2$fitted-mean(data2$Draft_No)))
B=2000
S_boot=rep(NA, B)
for (b in 1:B){
  data_boot=data2[sample(1:nrow(data2), replace = TRUE),]
  model_boot=loess(Draft_No ~ Day_of_year , data=data_boot)
  S_boot[b]=sum(abs(model_boot$fitted-mean(data_boot$Draft_No)))
}
p_value=sum(S_boot>S)/B
p_value

```

(c) Repeat Steps 5a--5b for k = 1, ... , until you have observed a couple of rejections.

```{r 1.5c}
#(c)

```

How good is your test statistic at rejecting the null hypothesis of a random lottery?

# Question 2: Bootstrap, jackknife and coincidence intervals

The data you are going to continue analyzing is the database of home prices in Albuquerque, 1993. The variables present are Price; SqFt: the area of a house; FEATS: number of features such as dishwasher, refrigerator and so on; Taxes: annual taxes paid for the house. Explore the file prices1.xls. The source of the original is the Data and Story Library (<https://dasl.datadescription.com/>) and it can be recovered from (<https://web.archive.org/web/20151022095618/http://lib.stat.cmu.edu/DASL/Datafiles/homedat.html>).

## 1.

Create a scatter plot of SqFt versus Price. Fit a linear model to it. Does a straight line seem like a good fit?

```{r 2.1}
data <- read.csv("prices1.csv", sep=";")

## 1)

price <- data$Price
sqft <- data$SqFt

plot(sqft, price)
fit1 <- lm(Price ~ SqFt, data)
lines(sqft, fit1$fitted.values, col="darkred", type="l", lwd=1)
```

A straight line seems to summarize the data pretty well.

## 2.

While the data do seem to follow a linear trend, a new sort of pattern seems to appear around 2000ft2. Consider a new linear model

$Price = b + a_1 * SqFt + a_2 * (SqFt - c) * 1_{SqFt>c}$ where c is the area value where the model changes. You can determine c using an optimizer, e.g., optim(), with the residual sum of squares (RSS) as the value to be minimized. For each value of c, the objective function should estimate b, a1, and a2; then calculate (and return) the resulting RSS.

```{r 2.2}
## 2)

fit2 <- function(c, data){
  data$new <- rep(0,110)
  for (i in 1:nrow(data)){
    if (data[i,2] > c) data[i,5] <- data[i,2] - c
    else data[i,5] <- 0
  }
  lm2 <- lm(Price ~ SqFt + new, data)
  sum(lm2$residuals^2)
}

RSS_opt <- optim(2000, fit2, data=data, method="L-BFGS-B", lower=0, upper=4500)
RSS_opt$par
RSS_opt$value
```

## 3.

Using the bootstrap estimate the distribution of c. Determine the bootstrap bias{correction and the variance of c. Compute a 95% confidence interval for c using bootstrap percentile, bootstrap BCa, and first{order normal approximation

(Hint: use boot(),boot.ci(),plot.boot(),print.bootci())

```{r 2.3}
library(boot)

c_val <- c()

for (i in 1:200){ #2000 takes too long to run on my poor laptop
  if (i%%500==0) cat(i, "\r")
  id <- sample(1:110, 110, replace = TRUE)
  data_boot <- data[id,]
  c_opt <- optim(2000, fit2, data=data_boot, method="L-BFGS-B", lower=0, upper=4500)   
  c_val <<- c(c_val, c_opt$par)
}

hist(c_val)

2 * RSS_opt$par - sum(c_val) / 2000      # bootstrap bias-correction
sum((c_val - mean(c_val))^2) / (2000-1)  # variance of c

opt_boot <- function(data, id){
  id <- sample(1:110, 110, replace = TRUE)
  data_boot2 <- data[id,]
  c_opt2 <- optim(2000, fit2, data= data_boot2, method = "L-BFGS-B", lower = 0, upper = 4500)   
  c_opt2$par
}

boot_fun <- boot(data = data, opt_boot, R = 2E3)
boot.ci(boot_fun)
plot(boot_fun)
```

We seem to obtain two Gaussian-ish shaped distributions as the distribution of the optimization values.

## 4.

Estimate the variance of c using the jackknife and compare it with the bootstrap estimate.

```{r 2.4}

jack_c <-c()

# new function for the jackknife function with 109 rows instead of 110
fit3 <- function(c, data){
  data$new <- rep(0,109)
  for (i in 1:nrow(data)){
    if (data[i,2] > c) data[i,5] <- data[i,2] - c
    else data[i,5] <- 0
  }
  lm2 <- lm(Price ~ SqFt + new, data)
  sum(lm2$residuals^2)
}

for (i in 1:110){
  id <- sample(1:110, 110, replace = TRUE)
  data_boot3 <- data[id,]
  data_jack <- data_boot3[-i,]
  c_opt <- optim(2000, fit3, data=data_jack, method = "L-BFGS-B", lower = 0, upper = 4500)   
  jack_c <<- c(jack_c, c_opt$par)
}

jack_c
var(jack_c)
```

(The values we obtain are not accurate, we will wait for feedback to understand what we did wrong.)

## 5.

Summarize the results of your investigation by comparing all of the confidence intervals with respect to their length and the location of c inside them.

# Appendix

```{r appendix, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
